Parallel-in-time methods for Particle Simulations

### 2024-12-02 paper/main.tex
Fixed bugs

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage[thinc]{esdiff}

@@

This is repeated until satisfactory convergence. It has been proven for example by \cite{parareal} (See Proposition 1) that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.

@@

@@

\subsection{Choice of the coarse and fine integrators}
\begin{itemize}
    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision {\color{red} TODO: source}. 
    \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic. A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
    \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method. However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
\end{itemize}

\subsection{Stopping criteria}
We assume the algorithm has converged when the maximum distance between the states of two consecutive iterations is smaller than some $\varepsilon$. We use $\varepsilon = 10^{-3}\text{ AU}$.

@@

\section{Implementation and experiments}
Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP.

@@

@@

\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator. 
$N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).
\newline The programs were run on the machines of the zia.cerit-sc.cz cluster owned by CERIT-SC/MU, a part of the MetaCentrum organization, each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\begin{center}
\begin{tabular}{ccrrrrrrrrcc}
\toprule
Program & Time &  n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs. error $^{\mathrm{b}}$ & Rel. error $^{\mathrm{c}}$ & CPU time & Runtime \\
\midrule
parareal & $10^4\text{days}$ & 32 & 32 &  $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
parareal & $10^4\text{days}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
serial & $10^4\text{days}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{b}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ Maximum relative error across the trajectory compared to the serial computation} \\
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}

@@

\begin{thebibliography}{00}
\bibitem{parareal}Bal, G. \& Maday, Y. A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put. {\em Recent Developments In Domain Decomposition Methods}. pp. 189-202 (2002)
6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), http://dx.doi.org/10.1086/118485
\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: http://publications.mfo.de/handle/mfo/3693
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: https://doi.org/10.1137/05064607X
\end{thebibliography}

>>>>

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{comment}
\usepackage[thinc]{esdiff}

@@

This is repeated until satisfactory convergence. It has been proven for example by \cite{parareal} (See Proposition 1) {\color{red} TODO: better source} that for $i \to \infty$, we will converge to $u_{0\dots N}$ such that $u_{n+1} = \mathcal{F}(t_n, t_n+\Delta T, u_n)$, which is equivalent to computing the states at times $t_{0\dots N}$ with the fine solver sequentially.

@@

\subsection*{Stopping criteria}
We assume the algorithm has converged enough when $\forall n: |u^{i+1}_n - u^i_n| < \varepsilon$. Other criteria also exist.\cite{speedup}

\subsection*{Theoretical speed-up}

A derivation of the theoretical speedup is presented in \cite{speedup} (Section 2.3), which we use here. It calculates the speed-up when running even the coarse solver partially in parallel, we simplify it with running $\mathcal{C}$ serially.

To find the theoretical speed-up of the algorithm, let's compute the time required by the parareal algorithm, $T_\text{parareal}$, and the time $T_\text{serial}$ taken to compute the solution by running the fine integrator serially.

Assume running $\mathcal{F}$ over any segment $[t_n, t_{n+1}]$ takes $T_\mathcal{F}$ and running $\mathcal{C}$ over any segment takes $T_\mathcal{C}$. Then computing $\mathcal{F}$ serially takes
$$
T_\text{serial} \approx NT_\mathcal{F}
$$
while using the parareal algorithm, with $N$ CPUs, we have
$$
T_\text{parareal} \approx NT_\mathcal{C} + \sum_{i=1}^{k}\left(T_\mathcal{F} + NT_\mathcal{C}\right) = kT_\mathcal{F} + (k+1)NT_\mathcal{C}
$$

Finally, the speed-up $S_\text{parareal}$ can be computed as
\begin{multline}
S_\text{parareal} = \frac{T_\text{serial}}{T_\text{parareal}} = \\\frac{NT_\mathcal{F}}{kT_\mathcal{F} + (k+1)NT_\mathcal{C}} = \\\left(\frac{k}{N} + (k+1)\frac{T_\mathcal{C}}{T_\mathcal{F}}\right)^{-1}
\end{multline}

Thus, to reach a high speed-up, we want the ratio $\frac{T_\mathcal{C}}{T_\mathcal{F}}$ to be small while also keeping the number of required iterations $k$ small.\cite{speedup}

The algorithm always converges in at most $N+1$ iterations, however, that is not useful (and takes even more time), because we have to converge at a significantly smaller number of iterations to gain any real speed-up.\cite{speedup}

@@

\subsection{Choice of the coarse and fine integrators}
\begin{itemize}
    \item We use the Velocity Verlet method as the coarse integrator for its favourable stability properties {\color{red} TODO: source} and the classic fourth-order Runge-Kutta method (RK4) as the fine integrator for its precision. 
    \item Two symplectic methods could be used, however, due to the nature of the parareal algorithm, the resulting parareal integration would not necessarily be symplectic. A modified parareal algorithm that preserves the symplectic property, shown in \cite{symplecticparareal}, also exists, however it is beyond the scope of this paper.
    \item An adaptive step size fine integrator with error control would be an option, such as the Dormand-Prince RK45 method. However, since the method can have very different execution times (depending on the step size required to reach desired accuracy), which could potentially be an issue when paralellizing the code, RK4 was used for simplicity.
\end{itemize}

@@

\section{Implementation and experiments}
Two implementations have been created, first a reference implementation in Python and then a faster one in C++ with OpenMP. {\color{red} TODO: rewrite}

@@

Let's derive the expected speed-up of our program and compare it with the measured speed-up. To simplify the calculation, we will only estimate it using the number of function evaluations. For Velocity Verlet, we have $T_\mathcal{C} = 1N_\mathcal{C}$, where $N_\mathcal{C}$ is the number of steps of the coarse integrator. RK4 takes four function evaluations, so we have $T_\mathcal{F} = 4N_\mathcal{F}$. If the fine integrator does 10 times more steps than the coarse one, we have $\frac{N_\mathcal{C}}{N_\mathcal{F}} = \frac{1}{10}$. Thus, we have
$$
S_\text{parareal} = \left(\frac{k}{N} + (k+1)\frac{1N_\mathcal{C}}{4N_\mathcal{F}}\right)^{-1} = \left(\frac{k}{N} + \frac{k+1}{40}\right)^{-1}
$$
When we choose the number of processors (and the number of segments), $N = 16$ and we converge within one iteration, we have
$$
S_\text{parareal} = \left(\frac{1}{16} + \frac{1}{20}\right)^{-1} \approx 8.89
$$

@@

\begin{comment}
\begin{table*}[htbp]
\caption{Comparison of running times taken by different configurations of the simulator. 
$N_{\text{Coarse}}$ and $N_{\text{Fine}}$ refer to the number of steps of the coarse and the fine integrators respectively. 
To allow comparison with the classic sequential algorithm using only the fine integrator, $N_{\text{Final}}$ shows the resulting number of steps (for parareal it is equal to the number of segments times $N_\text{Fine}$).}
\begin{center}
\begin{tabular}{ccrrrrrrrrcc}
\toprule
Program & Environment & n CPUs & Segments & $N_\text{Coarse}$ & $N_{\text{Fine}}$ & $N_{\text{Final}}$ & Iters $^{\mathrm{a}}$ & Abs. err. $^{\mathrm{b}}$ & Rel. err. $^{\mathrm{c}}$ & CPU time & Runtime \\
\midrule
\multicolumn{11}{c}{Solar system $10^4\text{days}$} \\
\midrule[.025em]
parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 32 & 32 & $5\times10^3$ &  $5\times10^4$ & $16\times10^5$ & 1 & 0.0030 AU & 0.97\% & 00:05:33 & 00:00:35 \\
parareal & zia.cerit-sc.cz$^{\mathrm{d}}$ & 16 & 16 & $1\times10^4$ & $1\times10^5$ & $16\times10^5$ & 2 & 0.0031 AU & 1.03\% & 00:08:12 & 00:01:07 \\
serial & zia.cerit-sc.cz$^{\mathrm{d}}$ & 1 & 1 & N/A & $16\times10^5$ & $16\times10^5$ & N/A & 0 AU & 0\% & 00:02:34 & 00:02:34 \\
\bottomrule
\multicolumn{11}{l}{$^{\mathrm{a}}$ Number of iterations required for convergence ($\varepsilon = 10^{-3}\text{ AU}$)} \\
\multicolumn{11}{l}{$^{\mathrm{b}}$ Maximum difference from the corresponding serial computation across the trajectory} \\
\multicolumn{11}{l}{$^{\mathrm{c}}$ Maximum relative error across the trajectory compared to the serial computation} \\
\multicolumn{11}{l}{$^{\mathrm{d}}$ The programs were run on the machines of the zia.cerit-sc.cz cluster } \\
\multicolumn{11}{l}{owned by CERIT-SC/MU, a part of the MetaCentrum organization,} \\
\multicolumn{11}{l}{each equiped with two AMD EPYC 7662 (2x 64 Core) 3.31 GHz processors.}
\end{tabular}
\label{runtimes}
\end{center}
\end{table*}
\end{comment}

@@

\begin{thebibliography}{00}
\bibitem{parareal}Bal, G. \& Maday, Y. A “Parareal” Time Discretization for Non-Linear PDE's with Application to the Pricing of an American Put. {\em Recent Developments In Domain Decomposition Methods}. pp. 189-202 (2002)
6-578 (2007), [Online]. Available: https://doi.org/10.1007/978-3-642-56118-4\_12
\bibitem{symplecticparareal}Bal, G. \& Wu, Q. Symplectic Parareal. {\em Domain Decomposition Methods In Science And Engineering XVII}. pp. 401-408 (2008)
\bibitem{parallelsolar}Saha, P., Stadel, J. \& Tremaine, S. A Parallel Integration Method for Solar System Dynamics. {\em The Astronomical Journal}. \textbf{114} pp. 409 (1997,7), http://dx.doi.org/10.1086/118485
\bibitem{speedup}Pentland, K., Tamborrino, M., Sullivan, T., Buchanan, J. \& Appel, L. GParareal: a time-parallel ODE solver using Gaussian process emulation. {\em Statistics And Computing}. 33, 23 (2022,12), [Online]. Available: https://doi.org/10.1007/s11222-022-10195-y
\bibitem{farfuture}Gander, M. Is it possible to predict the far future before the near future is known accurately?. {\em Snapshots Of Modern Mathematics From Oberwolfach; 2019}. pp. 21 (2019), [Online]. Available: http://publications.mfo.de/handle/mfo/3693
\bibitem{parareal2}Gander, M. \& Vandewalle, S. ``Analysis of the Parareal Time‐Parallel Time‐Integration Method.'' SIAM Journal On Scientific Computing. 29, 556-578 (2007), [Online]. Available: https://doi.org/10.1137/05064607X
\end{thebibliography}
